\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2017-01-24 nobuild/jss_revision/sparseHessianFD.Rnw
copied jss article to vignette
preparing final jss submission

\date{December 22, 2016}

@@

Let $x_1$ and $x_2$ be two subsets of elements of $x$.
Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
derivatives between all elements in $x_1$ and all elements in $x_2$.
Furthermore, define an indexing function Ind$(x_i)$ that returns the
indices in the parameter vector
$x$ that correspond to the elements in $x_i$, and $\hess{x_1,x_2}{}$
as the Hessian elements in rows Ind$(x_1)$ and columns Ind$(x_2)$, and
in rows Ind$(x_2)$ and columns Ind$(x_1)$ (from the symmetry of the
Hessian).

From the log posterior density in Equation~\ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
posterior density will be sparse.

@@

These ideas are illustrated in Figures~\ref{fig:graph1} and
\ref{fig:graph2}.  Figure~\ref{fig:graph1adj} shows the sparsity pattern of the lower triangle of a
Hessian as an adjacency matrix, and Figure~\ref{fig:graph1pic} is the
associated graph with a proper vertex coloring.  Every column (and
thus, every pair of columns) in Figure~\ref{fig:graph1adj} has a non-zero element
in row 7, so there are no non-empty intersection sets across the
columns.  All vertices are connected to each other in the
intersection graph (Figure~\ref{fig:graph1int}), which requires seven colors for a proper coloring.  Estimating a
sparse Hessian with this partitioning scheme would be no more
efficient than treating the Hessian as if it were dense.
\begin{figure}[ptb]
  \begin{subfigure}[b]{.32\textwidth}\centering
  \begin{tabular}{r|ccccccc}
   & 1&2&3&4&5&6&7\\
    \hline
    1& 1&&&&&&\\
    2&1&1&&&&&\\
    3&0&0&1&&&&\\
    4&0&0&1&1&&&\\
    5&0&0&0&0&1&&\\
    6&1&0&0&0&1&1&\\
    7&1&1&1&1&1&1&1
  \end{tabular}
  \caption{Adjacency matrix.}\label{fig:graph1adj}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=blue!20] (v3) {3};
\node at (180:1) [fill=red!20] (v4) {4};
\node at (240:1) [fill=blue!20] (v5) {5};
\node at (300:1) [fill=red!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)
(v3) -- (v7)  (v4) -- (v7)
(v5) -- (v7)  (v6) -- (v7)
(v1) -- (v2)  (v3) -- (v4)
(v5) -- (v6);
\end{tikzpicture}
}{} %% end toggle
\caption{Sparsity graph.}\label{fig:graph1pic}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=yellow!20] (v3) {3};
\node at (180:1) [fill=purple!20] (v4) {4};
\node at (240:1) [fill=brown!20] (v5) {5};
\node at (300:1) [fill=white!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)  (v3) -- (v7)
(v4) -- (v7)  (v5) -- (v7)  (v6) -- (v7)
(v1) -- (v6)  (v2) -- (v6)  (v3) -- (v6)
(v4) -- (v6)  (v5) -- (v6)
(v1) -- (v5)  (v2) -- (v5)  (v3) -- (v5)
(v4) -- (v5)
(v1) -- (v4)  (v2) -- (v4)  (v3) -- (v4)
(v1) -- (v3)  (v2) -- (v3)
(v1) -- (v2);
\end{tikzpicture}
}{} %% end toggle
\caption{Intersection graph.}\label{fig:graph1int}
\end{subfigure}
\caption{Unpermuted matrix.}\label{fig:graph1}
\end{figure}

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
<<results='asis',echo=FALSE>>=
arrange(tab4, M) %>%
xtable(digits=c(rep(0,4),rep(1,8))) %>%
    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
      include.rownames=FALSE, include.colnames=FALSE)
@
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
 difference and complex step methods, across 200 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}

>>>>

\date{January 25, 2017}

@@

Let $x_1$ and $x_2$ be two subsets of elements of $x$.
Define $\parD{2}{x_1,x_2}$ as the product set of cross-partial
derivatives between all elements in $x_1$ and all elements in
$x_2$. From the log posterior density in Equation~\ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.   Each $\parD{2}{}$ is
mapped to a submatrix of $\hess{}{}$, most of which with be zero.  The
resulting Hessian of the log posterior density will be sparse.

@@

These ideas are illustrated in Figures~\ref{fig:graph1} and
\ref{fig:graph2}.  Figure~\ref{fig:graph1adj} shows the sparsity pattern of the lower triangle of a
Hessian as an adjacency matrix, and Figure~\ref{fig:graph1pic} is the
associated graph with a proper vertex coloring.  Every column (and
thus, every pair of columns) in Figure~\ref{fig:graph1adj} has a non-zero element
in row 7, so there are no non-empty intersection sets across the
columns.  All vertices are connected to each other in the
intersection graph (Figure~\ref{fig:graph1int}), which requires seven colors for a proper coloring.  Estimating a
sparse Hessian with this partitioning scheme would be no more
efficient than treating the Hessian as if it were dense.
\begin{figure}[ptb]
  \begin{subfigure}[b]{.32\textwidth}\centering
  \begin{tabular}{r|ccccccc}
   & 1&2&3&4&5&6&7\\
    \hline
    1& 1&&&&&&\\
    2&1&1&&&&&\\
    3&0&0&1&&&&\\
    4&0&0&1&1&&&\\
    5&0&0&0&0&1&&\\
    6&0&0&0&0&1&1&\\
    7&1&1&1&1&1&1&1
  \end{tabular}
  \caption{Adjacency matrix.}\label{fig:graph1adj}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=blue!20] (v3) {3};
\node at (180:1) [fill=red!20] (v4) {4};
\node at (240:1) [fill=blue!20] (v5) {5};
\node at (300:1) [fill=red!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)
(v3) -- (v7)  (v4) -- (v7)
(v5) -- (v7)  (v6) -- (v7)
(v1) -- (v2)  (v3) -- (v4)
(v5) -- (v6);
\end{tikzpicture}
}{} %% end toggle
\caption{Sparsity graph.}\label{fig:graph1pic}
\end{subfigure}
\begin{subfigure}[b]{.32\textwidth}\centering
\iftoggle{tikz}{
\begin{tikzpicture}
\tikzstyle{every node}=[draw, shape=circle]
\node at (0,0) [fill=green!20] (v7) {7};
\node at (0:1) [fill=blue!20] (v1) {1};
\node at (60:1) [fill=red!20] (v2) {2};
\node at (120:1) [fill=yellow!20] (v3) {3};
\node at (180:1) [fill=purple!20] (v4) {4};
\node at (240:1) [fill=brown!20] (v5) {5};
\node at (300:1) [fill=white!20] (v6) {6};
\draw
(v1) -- (v7)  (v2) -- (v7)  (v3) -- (v7)
(v4) -- (v7)  (v5) -- (v7)  (v6) -- (v7)
(v1) -- (v6)  (v2) -- (v6)  (v3) -- (v6)
(v4) -- (v6)  (v5) -- (v6)
(v1) -- (v5)  (v2) -- (v5)  (v3) -- (v5)
(v4) -- (v5)
(v1) -- (v4)  (v2) -- (v4)  (v3) -- (v4)
(v1) -- (v3)  (v2) -- (v3)
(v1) -- (v2);
\end{tikzpicture}
}{} %% end toggle
\caption{Intersection graph.}\label{fig:graph1int}
\end{subfigure}
\caption{Unpermuted matrix.}\label{fig:graph1}
\end{figure}

@@

\begin{table}[ptb]\centering
  \begin{tabularx}{1\linewidth}{@{\extracolsep{\fill} }rrr|rrrr|rrrr}
  \toprule
&&&\multicolumn{4}{c|}{\bfseries{Finite differencing}}&\multicolumn{4}{c}{\bfseries{Complex step}}\\
&&&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c|}{\pkg{sparseHessianFD}}&\multicolumn{2}{c}{\pkg{numDeriv}}&\multicolumn{2}{c}{\pkg{sparseHessianFD}}\\
$N$&$k$&$M$&mean&sd&mean&sd&mean&sd&mean&sd\\
%%\midrule
<<results='asis',echo=FALSE>>=
arrange(tab4, M) %>%
xtable(digits=c(rep(0,4),rep(1,8))) %>%
    print(only.contents=TRUE, comment=FALSE, booktabs=TRUE,
      include.rownames=FALSE, include.colnames=FALSE)
@
\end{tabularx}
\caption{Computation times (milliseconds) for computing Hessians using the
  \pkg{numDeriv} and \pkg{sparseHessianFD} packages, and the finite
 difference and complex step methods, across 500 replications. Rows are ordered by the
  number of variables.}\label{tab:numDeriv}
\end{table}
